% Copyright (C) 2014-2024 by Thomas Auzinger <thomas@auzinger.name>

\documentclass[draft,final]{vutinfth} % Remove option 'final' to obtain debug information.

% Define convenience functions to use the author name and the thesis title in the PDF document properties.
\newcommand{\authorname}{Elias Huhsovitz} % The author name without titles.
\newcommand{\thesistitle}{Elastic Self-Adaptive Edge Pipelines} % The title of the thesis. The English version should be used, if it exists.

% Create the XMP metadata file for the creation of PDF/A compatible documents.
\begin{filecontents*}[overwrite]{\jobname.xmpdata}
\Author{\authorname}                                    % The author's name in the document properties.
\Title{\thesistitle}                                    % The document's title in the document properties.
\Language{de-AT}                                        % The document's language in the document properties. Select 'en-US', 'en-GB', or 'de-AT'.
\Keywords{a\sep list\sep of\sep keywords}               % The document's keywords in the document properties (separated by '\sep ').
\Publisher{TU Wien}                                     % The document's publisher in the document properties.
\Subject{Thesis}                                        % The document's subject in the document properties.
\end{filecontents*}

% Load packages to allow in- and output of non-ASCII characters.
\usepackage{lmodern}        % Use an extension of the original Computer Modern font to minimize the use of bitmapped letters.
\usepackage[T1]{fontenc}    % Determines font encoding of the output. Font packages have to be included before this line.
\usepackage[utf8]{inputenc} % Determines encoding of the input. All input files have to use UTF8 encoding.

% Extended LaTeX functionality is enables by including packages with \usepackage{...}.
\usepackage{amsmath}    % Extended typesetting of mathematical expression.
\usepackage{amssymb}    % Provides a multitude of mathematical symbols.
\usepackage{mathtools}  % Further extensions of mathematical typesetting.
\usepackage{microtype}  % Small-scale typographic enhancements.
\usepackage[inline]{enumitem} % User control over the layout of lists (itemize, enumerate, description).
\usepackage{multirow}   % Allows table elements to span several rows.
\usepackage{booktabs}   % Improves the typesetting of tables.
\usepackage{subcaption} % Allows the use of subfigures and enables their referencing.
\usepackage[ruled,linesnumbered,algochapter]{algorithm2e} % Enables the writing of pseudo code.
\usepackage[dvipsnames,table]{xcolor} % Allows the definition and use of colors. This package has to be included before tikz.
\usepackage{nag}        % Issues warnings when best practices in writing LaTeX documents are violated.
\usepackage{todonotes}  % Provides tooltip-like todo notes.
\usepackage{morewrites} % Increases the number of external files that can be used.
\usepackage[a-2b,mathxmp]{pdfx}      % Enables PDF/A compliance. Loads the package hyperref and has to be included second to last.
\usepackage[acronym,toc]{glossaries} % Enables the generation of glossaries and lists of acronyms. This package has to be included last.

% packages added by elias huhsovitz
\usepackage{float}

% used for tikz vector drawing
\usepackage{pgfplots}
\pgfplotsset{compat=1.18}

% Set PDF document properties
\hypersetup{
    pdfpagelayout   = TwoPageRight,           % How the document is shown in PDF viewers (optional).
    linkbordercolor = {Melon},                % The color of the borders of boxes around hyperlinks (optional).
}

\setpnumwidth{2.5em}        % Avoid overfull hboxes in the table of contents (see memoir manual).
\setsecnumdepth{subsection} % Enumerate subsections.

\nonzeroparskip             % Create space between paragraphs (optional).
\setlength{\parindent}{0pt} % Remove paragraph indentation (optional).

\makeindex      % Use an optional index.
\makeglossaries % Use an optional glossary.
%\glstocfalse   % Remove the glossaries from the table of contents.

% Set persons with 4 arguments:
%  {title before name}{name}{title after name}{gender}
%  where both titles are optional (i.e. can be given as empty brackets {}).
\setauthor{}{\authorname}{}{male}
% \setadvisor{Univ.Prof. Dr.}{Schahram Dustdar}{}{male}
\setadvisor{Proj.Ass. Dr.}{Boris Sedlak}{}{male}

% For bachelor and master theses:
% \setfirstassistant{Proj.Ass. Dr.}{Boris Sedlak}{}{male}
%\setsecondassistant{Pretitle}{Forename Surname}{Posttitle}{male}
%\setthirdassistant{Pretitle}{Forename Surname}{Posttitle}{male}

% For dissertations:
\setfirstreviewer{Pretitle}{Forename Surname}{Posttitle}{male}
\setsecondreviewer{Pretitle}{Forename Surname}{Posttitle}{male}

% For dissertations at the PhD School and optionally for dissertations:
\setsecondadvisor{Pretitle}{Forename Surname}{Posttitle}{male} % Comment to remove.

% Required data.
\setregnumber{12121272}
\setdate{01}{06}{2025} % Set date with 3 arguments: {day}{month}{year}.
\settitle{\thesistitle}{Active Inference for Elastic Control of Edge-Based Data Streams} % Sets English and German version of the title (both can be English or German). If your title contains commas, enclose it with additional curvy brackets (i.e., {{your title}}) or define it as a macro as done with \thesistitle.
%\setsubtitle{} % Sets English and German version of the subtitle (both can be English or German).

% Select the thesis type: bachelor / master / doctor.
% Bachelor:
\setthesis{bachelor}
%
% Master:
%\setthesis{master}
%\setmasterdegree{dipl.} % dipl. / rer.nat. / rer.soc.oec. / master
%
% Doctor:
%\setthesis{doctor}
%\setdoctordegree{rer.soc.oec.}% rer.nat. / techn. / rer.soc.oec.

% For bachelor and master:
\setcurriculum{Software \& Information Engineering}{Software \& Information Engineering} % Sets the English and German name of the curriculum.

% Optional reviewer data:
\setfirstreviewerdata{Affiliation, Country}
\setsecondreviewerdata{Affiliation, Country}


\begin{document}

\frontmatter % Switches to roman numbering.
% The structure of the thesis has to conform to the guidelines at
%  https://informatics.tuwien.ac.at/study-services

%\addtitlepage{naustrian} % German title page.
\addtitlepage{english} % English title page.
\addstatementpage

%\begin{danksagung*}
%\todo{Ihr Text hier.}
%\end{danksagung*}

\begin{acknowledgements*}
I would like to thank my supervisor, Dr. Boris Sedlak, for his guidance and helpful feedback during the development of this thesis.

I also thank Felix Berger and Tobias Seczer for their valuable non-technical feedback, particularly regarding figure layout and document presentation.
\end{acknowledgements*}

%\begin{kurzfassung}
%\todo{Ihr Text hier.}
%\end{kurzfassung}

\begin{abstract}
This thesis explores how the Active Inference (AIF) framework can be employed to uphold Service Level Objectives (SLOs) in resource-constrained edge computing environments. A novel distributed stream processing pipeline is designed and implemented, enabling real-time video inference across parallel worker nodes. Elasticity is expressed through dynamic adjustment of three stream quality parameters: frame rate, resolution, and model complexity.

An AIF agent deployed at the producer node continuously monitors the system state and minimizes expected free energy to select actions that reduce the risk of future SLO violations. Online learning is achieved through runtime updates to the agent's generative model. SLOs—such as task queue size, processing latency, and memory usage—are encoded as prior preferences, allowing the agent to balance constraint satisfaction with stream quality maximization.

The prototype is evaluated against a heuristic baseline across three scenarios: a static environment, dynamic demand variation, and fluctuating computational budget. Results show that the AIF agent outperforms the baseline in stable and moderately dynamic environments, maintaining higher global SLO satisfaction and more stable configurations. However, performance degrades under highly volatile workloads due to limited reactivity. This highlights the need for future extensions supporting faster re-adaptation and distributed control across multiple nodes.

Overall, this work presents a principled method for elastic, self-adaptive control in edge systems. By integrating probabilistic reasoning, preference-driven adaptation, and online model learning, Active Inference provides a unified control paradigm for resilient stream processing under uncertainty.
\end{abstract}


% Select the language of the thesis, e.g., english or naustrian.
\selectlanguage{english}

% Add a table of contents (toc).
\tableofcontents % Starred version, i.e., \tableofcontents*, removes the self-entry.

% Switch to arabic numbering and start the enumeration of chapters in the table of content.
\mainmatter

% link to chapters
\include{chapters/introduction}
\include{chapters/background}
\include{chapters/related_work}
\include{chapters/methodology}
\include{chapters/implementation}
\include{chapters/evaluation}
\include{chapters/discussion}
\include{chapters/conclusion}


% Remove following line for the final thesis.
%\input{intro.tex} % A short introduction to LaTeX.

\backmatter

% Declare the use of AI tools as mentioned in the statement of originality.
% Use either the English aitools or the German kitools.
\begin{aitools}

In the course of writing this thesis, several generative AI-based tools were used to support specific tasks related to grammar correction, LaTeX troubleshooting, and literature note-taking. The tools listed below were used selectively and within the scope permitted for academic writing support.

\begin{itemize}
    \item \textbf{Grammarly} -- Used primarily for spell and grammar checking of written text, limited to basic proofreading. Available at \url{https://app.grammarly.com/}.
    
    \item \textbf{Writefull (Overleaf Integration)} -- Integrated into Overleaf to provide in-editor grammar and language feedback. Used solely to identify minor writing issues. More information at \url{https://www.writefull.com/}.
    
    \item \textbf{Overleaf AI Assist} -- Employed to resolve LaTeX syntax and compilation errors during document preparation. Details can be found at \url{https://www.overleaf.com/learn/how-to/AI_Assist}.
    
    \item \textbf{Google NotebookLM} -- Used as a digital notebook for organizing literature notes and retrieving specific excerpts from uploaded documents. Accessible at \url{https://notebooklm.google.com/}.
\end{itemize}

All generative AI tools were used transparently, with careful attention to maintaining the integrity and originality of the thesis content.
\end{aitools}

% Use an optional list of figures.
\listoffigures % Starred version, i.e., \listoffigures*, removes the toc entry.

% Use an optional list of tables.
\cleardoublepage % Start list of tables on the next empty right hand page.
\listoftables % Starred version, i.e., \listoftables*, removes the toc entry.

% Use an optional list of alogrithms.
\listofalgorithms
\addcontentsline{toc}{chapter}{List of Algorithms}

% Add an index.
\printindex

% Add a glossary.
\printglossaries

% Add a bibliography.
\bibliographystyle{ieeetr}
\bibliography{bib}

\appendix
\include{appendix/simulation_results}

\end{document}