\chapter{Related Work}
\label{chap:related-work}
This chapter provides an overview of key literature relevant to Elasticity and AIF within distributed and edge computing contexts. While AIF has found wide application in robotics and cognitive systems~\cite{lanillos_active_2021}, its adoption in edge computing remains sparse. Notable progress in applying elasticity through AIF in resource-constrained environments emerged primarily through the work of Sedlak et al.~\cite{sedlak_adaptive_2024, sedlak_equilibrium_2024, sedlak_active_2024, sedlak_slo-aware_2025, lapkovskis_benchmarking_2025, sedlak_towards_2025}, Danilenka et al.~\cite{danilenka_adaptive_2025}, and to a certain extent, Levchuk et al.~\cite{levchuk_active_2019} in multi-agent Internet
of Things (IoT) systems. Given the early stage of AIF in edge computing, this chapter focuses on a selected set of foundational and state-of-the-art contributions to contextualize the motivation and relevance of our approach.

\section{Runtime adaptation of stream processing systems}
Runtime adaptation allows stream processing services to continuously adapt to changing workloads and environments. Cardellini et al.~\cite{cardellini_runtime_2022} review a wide range of techniques that systems use to adapt at runtime. These include resource scaling, relocating operators across devices, skipping or dropping data under overload (load shedding), and fine-tuning system parameters like buffer size or batch interval. Adaptation can be managed centrally by a global controller or locally by decentralized agents, and can be triggered either in reaction to performance violations or in anticipation of them.

Early implementations such as DivProg by Fürst et al.~\cite{furst_elastic_2018} demonstrate heuristic-based adaptation using function annotations and a code selection engine. Their elastic services dynamically switch between classifier types and image resolutions to meet latency constraints across heterogeneous edge devices.

Cao et al.~\cite{cao_neural_2023} present RL-Adapt, a framework that trains reinforcement learning models to generate adaptation policies targeting application-specific QoE metrics. Their system operates in real time on constrained IoT hardware, adjusting parameters like frame rate and compression quality.

\section{Elasticity for stream processing under SLO constraints}
In~\cite{sedlak_towards_2025}, Sedlak et al. extend this approach, presenting a control architecture where services can adapt along multiple dimensions of elasticity, i.e., resource- and quality-elasticity. The system dynaimcally adjusts configuration parameters such as the number of allocated CPU cores (\textit{cpu}) and the resolution of the video stream (\textit{pixels}). These two controllable parameters, along with the frame rate (\textit{fps}) of the video stream, are used as SLO targets for the stream processing services. Actions are selected based on a combination of Deep Reinforcement Learning and a reward function, based on SLO fulfillment. Rather than relying solely on vertical autoscaling, their framework uses elasticity as a first-class mechanism to operate under fixed-resource budgets typical of edge nodes. This work also introduces the idea of causally linking low-level service parameters with high-level application goals.

\section{Active Inference as a unified approach for edge adaptation}
In~\cite{sedlak_active_2024}, Sedlak et al. present a design study for applying Active Inference (AIF) to edge computing systems with a focus on maintaining Service Level Objectives (SLOs) in resource-constrained environments. They propose a self-evidencing agent that operates in an action-perception cycle, using a generative model to predict system behavior, compare outcomes with expectations, and update beliefs to minimize free energy. The agent’s preferences are encoded directly as SLOs, guiding both its predictions and actions. The framework explicitly links system metrics, such as batch delay and utilization, to causal structures. This enables interpretability and continuous adaptation without requiring pretraining. This approach distinguishes itself from traditional ML systems by integrating causal reasoning, homeostasis, and epistemic exploration into a unified control loop. This design study provides the blueprint for implementing AIF-based control in edge computing scenarios.

Complementing this theoretical design,~\cite{sedlak_adaptive_2024} presents an implemented and evaluated AIF agent for adaptive stream processing on edge devices. Their system dynamically adjusts stream quality parameters---specifically frame rate (fps), resolution (pixel), and processing mode—based on the agent’s generative model and its beliefs about system behavior. SLOs such as latency, energy consumption, and detection accuracy are encoded as preferred observations (priors), guiding the agent’s policy selection through the minimization of EFE. The action-perception cycle allows the agent to iteratively update its beliefs and reconfigure services in response to observed deviations, thereby ensuring consistent SLO fulfillment across heterogeneous devices and workloads. This work directly influences the methodology of this thesis, which adopts a comparable architectural approach and operationalizes SLOs in a similar way to enable self-adaptive control.


